The Current State of R Tools for Nonlinear Least Squares Modeling

### 2023-07-01 RefactoringNLS/revision202307/ReviewResponses.txt

Review 1

Referee report of “A Comparison of R Tools for Nonlinear Least
Squares Modeling”.
ID: 2023-15
General
This is a long paper that covers many topics related to nonlinear least squares fitting with R. This
mixture of many topics can be confusing and I think the paper would benefit from a clearer definition
of its purpose and focus. Once a clearer focus is established it will be easier to decide which topics are
relevant to that purpose.
The paper’s title indicates that the purpose is to compare and contrast facilities in base R, meaning
the nls function and related facilities in the stats package, and several external packages for nonlinear
least-squares modeling that are available on CRAN. There is a listing of packages related to nonlinear
least squares on page 2 but the majority of the discussion is focussed on nls, in the stats package that
is part of base R, and the nlsr package (Nash and Murdoch 2023), written by this paper’s authors and
others (on CRAN this package is said to be the work of four authors but only two of them are cited in
the paper).
There are also indications that the paper is about replacing nls in base R with functions from nlsr. The
abstract begins by citing the Google Summer of Code project “Improvements to nls()” and the first
sentence in the “Principal messages”” section indicates that the purpose is to create a refactored nls()
based on the work described here. Toward the end of the paper, the section on “Future of nonlinear
model estimation in R” also indicates that the goal is to enhance or replace nls() and related parts
of base R.
So is the purpose of the paper:
• To introduce the nlsr package and describe its capabilities?
• To compare and contrast several available packages for nonlinear least-squares?
• To advocate that the authors’ code replace nls in base R?
I would recommend concentrating on the first objective — introducing the nlsr package and demonstrating
the advantages that it provides relative to nls — and I will address my comments to that purpose. I
hope this approach will be of use to the authors and the editor.
It is certainly appropriate when describing the nlsr package to contrast it with facilities like nls() in
base R. To help with this I will provide some background on the development of nls() and the stats
package in general.
Detailed comments
nls is S and in R
Although nls() has been part of R for over 25 years, it originated much earlier in the “Statistical Models
in S” project, documented in Chambers and Hastie (1992), and based on work starting in the mid-1980’s.
The purpose was to create a unified set of tools for fitting statistical models, including nonlinear models
with a scalar objective, such as a log-likelihood, or as a nonlinear regression model. Many capabilities for
R, including S3 classes and methods, the formula language for linear predictors, and model frames and
model matrices, were based directly on the results of this project in S. (As described by Ross Ihaka, one
of the two original developers of R, their goal was to produce a language that was “not unlike S” so the
resemblance was not accidental.)
1What was called QPE (quantitative programming environment), then “The New S Language”, then “S3”
represented a huge step forward for nonlinear modeling because it could represent model functions as
expressions (i.e. abstract syntax trees) that could be manipulated in the system itself. This allowed for
symbolic differentiation.
The original implementation of the deriv function for symbolic differentiation was written in S. When
porting deriv to R in the mid-1990’s Ross Ihaka moved quite a bit of the functionality to C code because
it was comparatively easy to do so and, on the hardware of the day, the C implementation provided a
considerable speed advantage. The speed of the differentiation itself isn’t usually a problem - it’s the
common-sub-expression elimination that can be the bottleneck.
This legacy is also the reason that the Jacobian matrix is the "gradient" attribute of the returned value
of the model function. For a nonlinear least squares problem, where the model function returns a vector,
this deriv attribute should be called "Jacobian" but the version for the scalar objective, for which the
vector of partial derivatives is the gradient, was written first. Because deriv only has access to the
expression and not to the model frame it is not possible to determine if the value will be a scalar or a
vector and the name of the attribute was left as "gradient". A further complication is that numerical
values in S and R are always stored as vectors, with the understanding that a scalar is represented as
a vector of length 1. Even if it were desirable to distinguish between scalars and vectors, internally it
is impossible to determine if a vector of length 1 is intended to be a scalar or to be a vector that just
happens to have length 1.
The legacy is also the reason that initial data manipulation, such as evaluating a subset or na.action, is
performed by non-standard evaluation of a model.frame. The non-standard evaluation means that, within
the nls function evaluation, the code accesses its own call (with match.call) and re-writes the function
name to be stats::model.frame then evaluates the rewritten call (around line 580 of nls.R). Handling
all the special cases is complicated but the purpose is so that arguments like subset or na.action and
conversion of formulas to model frames to model matrices are handled consistently across the modeling
functions in the stats package and without code duplication.
Thomas Lumley has called this “replace the function name in my own call” approach the “standard,
non-standard evaluation idiom” in R.
Another aspect of this legacy is the expected behavior of print, summary, plot, etc. applied to an object
representing a fitted model. Peculiarities like summary producing longer descriptions than print or the
choice of printing the sum of squared residuals or the estimated standard deviation of the residuals are
for consistency with the other model output.
It is also why the extractor for the parameter values is coef, even though the parameters in a nonlin-
ear regression model are not necessarily coefficients, and why they are shown in the summary using
print.coefmat.
These choices, which allow users to transfer experience from one modelling situation to another, are
neither arbitrary nor trivial.
Function closure with a shared environment
One of the major differences between the S and R implementations of nls is the use of function closures
in R to allow several functions in the m field of an nls model (see below for an example) shared access to
parameter values, covariate values, the response vector and various pieces of auxillary information.
In R if a function is defined during the evaluation of another function its “closure” includes the evaluation
environment of its parent. If several functions are defined and returned, say in a list, then they have
access to this common environment. This shared environment is where parameter updates and subsequent
evaluation of the model function, the Jacobian and various decompositions occur.
Self-starting models in S and R
Self-starting models were developed for nlme package for S, Pinheiro and Bates (2000), to automate
the fitting of nonlinear regression models to repeated measures data collected on several observational
2units, e.g. several different subjects. Rather than requiring the analyst to provide starting values for each
subject’s data, the steps of formulating and refining such starting values were encapsulated in functions.
These functions also used symbolic derivatives to provide Jacobian evaluations and frequently fit a reduced
form of the model using the plinear algorithm to refine the starting values, resulting in the “starting
values” being the actual parameter estimates. These are fed into the default algorithm to produce a
summary but with the understanding that the default algorithm will converge at the first evaluation
(i.e. 0 iterative steps) if the convergence criterion is the relative offset. The point here is that the relative
offset criterion depends only on the current parameter values and not on changes in the parameter values
so there is only one additional evaluation required after the plinear algorithm converges.
For the illustrative example starting at the bottom of page 2 of this paper the SSlogis self-starting
model can be used.
weeddf <- data.frame(
tt = 1:12,
weed = c(
5.308, 7.24, 9.638, 12.866, 17.069, 23.192,
31.443, 38.558, 50.156, 62.948, 75.995, 91.972
)
)
summary(hobbsm1 <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weeddf))
Formula: weed ~ SSlogis(tt, Asym, xmid, scal)
Parameters:
Estimate Std. Error t value Pr(>|t|)
Asym 196.1862
11.3069
17.35 3.17e-08 ***
xmid 12.4173
0.3346
37.11 3.72e-11 ***
scal
3.1891
0.0698
45.69 5.77e-12 ***
---
Signif. codes: 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
Residual standard error: 0.5362 on 9 degrees of freedom
Number of iterations to convergence: 0
Achieved convergence tolerance: 1.118e-06
str(hobbsm1)
List of 6
$ m
:List of 16
..$ resid
:function ()
..$ fitted
:function ()
..$ formula
:function ()
..$ deviance :function ()
..$ lhs
:function ()
..$ gradient :function ()
..$ conv
:function ()
..$ incr
:function ()
..$ setVarying:function (vary = rep_len(TRUE, np))
..$ setPars
:function (newPars)
..$ getPars
:function ()
..$ getAllPars:function ()
..$ getEnv
:function ()
3..$ trace
:function ()
..$ Rmat
:function ()
..$ predict
:function (newdata = list(), qr = FALSE)
..- attr(*, "class")= chr "nlsModel"
$ convInfo
:List of 5
..$ isConv
: logi TRUE
..$ finIter
: int 0
..$ finTol
: num 1.12e-06
..$ stopCode
: int 0
..$ stopMessage: chr "converged"
$ data
: symbol weeddf
$ call
: language nls(formula = weed ~ SSlogis(tt, Asym, xmid, scal), data = weeddf, algorithm
$ dataClasses: Named chr "numeric"
..- attr(*, "names")= chr "tt"
$ control
:List of 7
..$ maxiter
: num 50
..$ tol
: num 1e-05
..$ minFactor : num 0.000977
..$ printEval : logi FALSE
..$ warnOnly
: logi FALSE
..$ scaleOffset: num 0
..$ nDcentral : logi FALSE
- attr(*, "class")= chr "nls"
This only takes a few milliseconds on a laptop computer
microbenchmark::microbenchmark(nls(weed ~ SSlogis(tt, Asym, xmid, scal), weeddf), times=100L)
Unit: milliseconds
expr
min
lq
mean
nls(weed ~ SSlogis(tt, Asym, xmid, scal), weeddf) 3.483984 3.624459 4.599596
median
uq
max neval
3.92275 5.035838 13.76178
100
although it turns out that the SSlogisJN formulation is slightly faster in this case
summary(hobbsm2 <- nls(weed ~ nlsr::SSlogisJN(tt, Asym, xmid, scal), weeddf))
Formula: weed ~ nlsr::SSlogisJN(tt, Asym, xmid, scal)
Parameters:
Estimate Std. Error t value Pr(>|t|)
Asym 196.1862
11.3069
17.35 3.17e-08 ***
xmid 12.4173
0.3346
37.11 3.72e-11 ***
scal
3.1891
0.0698
45.69 5.77e-12 ***
---
Signif. codes: 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
Residual standard error: 0.5362 on 9 degrees of freedom
Number of iterations to convergence: 3
Achieved convergence tolerance: 2.419e-06
microbenchmark::microbenchmark(nls(weed ~ nlsr::SSlogisJN(tt, Asym, xmid, scal), weeddf), times=10
4Unit: milliseconds
expr
min
lq
nls(weed ~ nlsr::SSlogisJN(tt, Asym, xmid, scal), weeddf) 1.960326 2.069342
mean
median
uq
max neval
2.492946 2.184441 2.554391 11.43154
100
Some of the discussion in the “Jacobian code in selfStart models” section (p. 9) of this paper claims
that the plinear fit followed by a fit using the default approach would be difficult to maintain and could
be very slow but that has not been the case. It is known (and, in fact, stated in the documentation for
SSlogis) that the call to the default nls algorithm will always declare convergence immediately.
Also in that section it is claimed that deriv creates “hidden variables” because the default for the tag
argument to deriv is .expr. This tag is only used in bindings within the evaluation environment of the
model function, which is never visible to the user unless they are inspecting the function evaluation with
a symbolic debugger. The only reason for the default tag starting with "." is to avoid conflicts with
parameter or covariate names in the model expression. But it is only the default.
Recommendations
As stated in the “General” section, I think the purpose of the paper should be clarified. If the purpose
is to introduce the nlsr package and contrast it with the existing implementation of nls in the stats
package, then it might be helpful to understand why nls was implemented the way it was.
I hope that this unconventional referee’s report is of use to the authors and the editor.
References
Chambers, John M., and Trevor J. Hastie, eds. 1992. Statistical Models in S. Routledge. https:
//doi.org/10.1201/9780203738535.
Nash, John C, and Duncan Murdoch. 2023. Nlsr: Functions for Nonlinear Least Squares Solutions -
Updated 2022. https://CRAN.R-project.org/package=nlsr.
Pinheiro, José, and Douglas Bates. 2000. Mixed-Effects Models in S and S-PLUS. Springer New York.
5
\

Review 2

Referee report: A Comparison of R Tools for Nonlinear Least
Squares Modeling
Id: 2023-15
2023-06-28
General

This article provides a vignette-style overview of the nlsr-package and highlights its design choices and
behavior in comparison to base R’s nls() function as well as the nonlinear least squares solvers provided by
the R-packages minpack.lm and gslnls. The article’s main contribution is to make R users (more) aware of
the historic shortcomings and pitfalls in using R’s nls() function and how the nlsr-package is able to serve
as a more reliable alternative. Overall, the article is well-written and structured into coherent subthemes
and sections. Given that nls() is the de facto standard for solving nonlinear least squares problems in R, I
believe this article provides valuable insights and is well-suited for publication in the R Journal after some
revisions and minor clarifications.

My main advice is to revise the title and introduction of the article. As it currently stands, it is suggested
that the article provides an (objective) comparison of nonlinear least squares solvers available in R. The body
of the article reads more like a vignette for the nlsr-package, explaining its design choices and comparing its
behavior to nls() and other NLS solvers based on a few selected examples. As such, a more appropriate title
could be, for instance, The current state of R tools for nonlinear least squares modeling or similar.

2:01

We appreciate the positive comments and agree with the suggested title change.
Both referees have requested revisions to the introduction, and we hope that we have addressed the
comments.

Detailed comments
1. The Hobbs weed example as an illustration of the performance of different NLS solvers is lacking a
thorough comparison. I understand that the aim of the authors is to provide a data example that
demonstrates the reliability of nlsr in favor of nls(), but nls() converges to the same solution as
nlsr for all model setups when using the Port algorithm. This goes unnoticed in the text. Also,
the model setup (Logistic3T) for which gsl_nls() fails can easily be made to converge correctly by
changing either the algorithm or the scaling method, see the Illustrating code section. My advice
is to provide a more comprehensive comparison of the available NLS solvers, or to pick a different
data example that better highlights the benefits of nlsr. For instance, trying to solve the BoxBOD
regression problem (https://www.itl.nist.gov/div898/strd/nls/data/boxbod.shtml), nls() fails with
starting values θ1 = 1, θ2 = 1 using both the Gauss-Newton and Port algorithm, whereas nlsr::nlxb()
converges correctly.
2. The section Returned results of nls() and other tools discusses the complexity of S3-objects of class
nls, illustrated by the retrieval of the input data from an nls-object. It seems ill-advised to run
eval(parse(text=result$data)), since it evaluates the name of the data object in the user’s envi-
ronment, which may not exist. Instead, one should use model = TRUE to store the input data in the
returned nls-object. This behavior is consistent with lm(), so perhaps less confusing than currently
presented in the manuscript. The authors mention that nlxb() returns a much simpler structure of 11
items in one level, but I did not manage to find the input data object in the nlsr-object returned by
nlxb().
3. The section Functional specification of problems mentions the use of nlsr::nlfb() to solve nonlinear
least squares problems using a function to define the residuals. Since nlsr::nlfb() only has a notion
of the residuals, how is it able to disentangle model and response, for the purpose of e.g. generating
1model predictions? The authors mention that the function wrapnlsr() can be used to return an object
of class nls, but this does not seem to work for problems defined through a residual function as in
nlsr::nlfb()? If indeed the case, how does one evaluate e.g. model predictions, parameter confidence
intervals, etc. when using nlsr::nlfb()? This remains unclear to me.
4. The section Philosophical considerations hints at the challenge of performing inference when parameter
bound-constraints are involved in the optimization. In case of physical or model constraints on the
parameters, a common approach is to reparametrize the model to be unconstrained in the parameters.
Perhaps the authors can highlight an example scenario where imposing parameter constraints is
preferable to model reparametrization (the included data examples with parameter constraints appear
somewhat artificial), or elaborate further on the possible downsides of reparametrizing the model, if any.
5. The section Programming language makes the argument that keeping to a single programming language
can allow for easier maintenance and upgrades, and that the performance penalty for using code entirely
in R has been much reduced in recent years. I agree that there is not much of a performance penalty
when solving small problems, such as the data examples included in the article. However, when solving
problems using a large amount of data or a large number of parameters, I have experienced that the
run-time (as well as memory usage) can become an actual bottleneck. As an example, consider the
Penalty function I problem (Moré, Garbow, and Hillstrom (1981)) included in the Illustrating code
section.
Minor comments
• Avoid emphasizing words by writing in all capitals. Examples: NOT (pg. 14, 15), WEIGHTED (pg.
13) and more.
• The link https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211 (as well as the links to github)
may be subject to change or become unavailable in the future.
Illustrating code
## Hobbs weed example
weeddf <- data.frame(
tt = 1:12,
weed = c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192,
31.443, 38.558, 50.156, 62.948, 75.995, 91.972)
)
### nls-port all setups
nls(
formula = weed ~ b1 / (1 + b2 * exp(-b3 * tt)),
data = weeddf,
start = c(b1 = 1, b2 = 1, b3 = 1),
algorithm = "port"
)
#> Nonlinear regression model
#>
model: weed ~ b1/(1 + b2 * exp(-b3 * tt))
#>
data: weeddf
#>
b1
b2
b3
#> 196.1863 49.0916
0.3136
#> residual sum-of-squares: 2.587
#>
#> Algorithm "port", convergence message: relative convergence (4)
nls(
formula = weed ~ 100 * c1 / (1 + 10 * c2 * exp(-0.1 * c3 * tt)),
2data = weeddf,
start = c(c1 = 1, c2 = 1, c3 = 1),
algorithm = "port"
)
#> Nonlinear regression model
#>
model: weed ~ 100 * c1/(1 + 10 * c2 * exp(-0.1 * c3 * tt))
#>
data: weeddf
#>
c1
c2
c3
#> 1.962 4.909 3.136
#> residual sum-of-squares: 2.587
#>
#> Algorithm "port", convergence message: relative convergence (4)
nls(
formula = weed ~ Asym /(1 + exp((xmid - tt) / scal)),
data = weeddf,
start = c(Asym = 1, xmid = 1, scal = 1),
algorithm = "port"
)
#> Nonlinear regression model
#>
model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
#>
data: weeddf
#>
Asym
xmid
scal
#> 196.186 12.417
3.189
#> residual sum-of-squares: 2.587
#>
#> Algorithm "port", convergence message: relative convergence (4)
### gsl_nls Logistic3T setup
gslnls::gsl_nls(
fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
data = weeddf,
start = c(Asym = 1, xmid = 1, scal = 1),
algorithm = "lmaccel"
)
#> Nonlinear regression model
#>

### 2023-07-03 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
Responses -- first draft

---
title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
author:
- John C. Nash \thanks{ retired professor, Telfer School of Management, University
  of Ottawa}
- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
  Institute of Technology, Kanpur}
date: "2023-3-2"
output:
  pdf_document:
    keep_tex: yes
    toc: no
  word_document:
    toc: no
bibliography: ./ImproveNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

Our Google Summer of Code project "Improvements to `nls()`" considered the features and limitations of the R function `nls()` with the aim 
of improving and rationalizing R tools for nonlinear regression.
The rich features of `nls()` are weakened by several deficiencies and
inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
Further considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Various packages, including our `nlsr`, provide alternative capabilities.
We consider the differences in goals, approaches, and features of
different tools for nonlinear least squares modeling in R. Discussion of 
these matters is relevant to improving R generally as well as its 
nonlinear estimation tools.

@@

`nls()` is the tool for estimating nonlinear statistical models in base R, the 
primary software distribution
from the Comprehensive R Archive Network (*https://cran.r-project.org*).
The function dates to the 1980s and the work related to @bateswatts in S (see 
*https://en.wikipedia.org/wiki/S_%28programming_language%29*).

@@

The tools we will consider are the base-R `nls()` function 
and some packages that are available on the CRAN repository.
Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
general nonlinear least-squares solvers.

While we will provide capsule comments for some other CRAN packages, 
we will not pursue those in the Bioconductor (@Bioconductor) 
collection, nor those on repositories such as 
GitHub (https://github.com) and Gitlab (https://about.gitlab.com).

@@

@@

`optimx` (@p-optimx) -- This provides 
optimizers that can be applied to minimize a nonlinear function which could be
a nonlinear sum of 
squares. NOT generally recommended if nonlinear 
least squares programs can be easily used, but provides a check and alternative
solvers.

@@

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by NOT forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods.

@@

Let us specify in R the three model formulas and set some starting
values for parameters. These starting points are NOT equivalent and
are deliberately crude choices. Workers performing many calculations 
of a similar nature should try to provide good starting points to reduce
computation time and avoid finding a false solution.

@@

@@

Though we have found solutions, the Jacobian is essentially singular as 
shown by its singular
values. Note that these are **displayed** by package `nlsr` in a single column in
the output to provide a compact layout, but the values do **NOT** correspond to 
the individual parameters in 
whose row they appear; they are a property of the whole problem.

@@

```{r ex02gslnls}
library(gslnls)
ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
pnls(ugslnls1) # to get sum of squares
sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
pnls(sgslnls1) # Use summary() to get display
tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
pnls(tgslnls1) 
```

@@

`nlsr::nlxb()` uses `print()` to output standard errors and 
singular values of the Jacobian (for diagnostic purposes). By contrast, 
`minpack.lm::nlsLM()` and `nls()`
use `summary()`, which does NOT display the sum of squares, while `print()` 
gives the sum of squares, but not the standard error of the residuals.

The singular values allow us to gauge how "nearly singular" the 
Jacobian is at the solution, and the ratio of the smallest to largest of the 
singular values is a simple but effective measure. The ratios are
4.6641e-05 for Logistic3U, 0.021022 for Logistic3S, and 0.001055 for
Logistic3T, so Logistic3S is the "least singular".

@@

We see that there are differences in detail, but the more important result is that
two out of three singular values are essentially 0. Our Jacobian is singular, and no 
method of the Gauss-Newton type should be able to continue. Indeed, from the parameters
reported at this saddle point, `nlsr::nlxb()` cannot proceed.

@@

@@

@@

We are also surprised that the 
analytic expressions for the Jacobian ("gradient") in the `SSLogis` function 
and others save quantities in "hidden"
variables, i.e., with names preceded by ".". 
These are then not displayed by the `ls()`
command, making them difficult to access by users who may wish to create
their own selfStart model via copy and edit.
Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
worth considering as another way to find starting parameters, but we know of no
R capability of this type.

@@

@@

`nls()` can be called without specifying the `data` argument. In this case, it will
search in the available environments (i.e., workspaces) for suitable data objects. 
We do NOT like this approach, but it is "the R way". R allows users to leave many 
objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
suppress saving this workspace (`.RData`) on exit; otherwise, any such file in 
the path will be loaded on startup. R users in our
acquaintance avoid saving the workspace because of lurking data and
functions that may cause unwanted results.

>>>>

---
title: "The Current State of R Tools for Nonlinear Least Squares Modeling"
author:
- John C. Nash \thanks{ retired professor, Telfer School of Management, University
  of Ottawa}
- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
  Institute of Technology, Kanpur}
date: "2023-3-2"
output:
  pdf_document:
    keep_tex: yes
    toc: no
  word_document:
    toc: no
bibliography: ./StateOfRToolsForNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

@@

Our Google Summer of Code project "Improvements to `nls()`" considered the features 
and limitations of the R function `nls()` with the aim 
of improving and rationalizing R tools for nonlinear regression.
The rich features of `nls()` are weakened by several deficiencies and
inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
Further considerations are the usability and  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Various packages, in particular our `nlsr`, provide alternative capabilities.
We consider the differences in goals, approaches, and features of
different tools for nonlinear least squares modeling in R. Discussion of 
these matters is relevant to improving R generally as well as its 
nonlinear estimation tools.

@@

`nls()` is the tool for estimating nonlinear statistical models in base R, the 
primary software distribution
from the Comprehensive R Archive Network (*https://cran.r-project.org*), hereafter
abbreviated CRAN.
Users do not have to request that `nls()` be loaded.
The function dates to the 1980s and the work related to @bateswatts and others 
in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).

@@

Our Google Summer of Code project for 2021, "Improvements to `nls()`"
had the goal of providing a version of the code that would address some
of the deficiencies discovered over the long period that the `nls()`
has been part of R. However, it was also critical to preserve legacy
functionality, and we did not achieve a result that provided desired
improvements and preserved legacy results. On the other hand, working
with `nls()` suggested ways our own `nlsr` package could be improved
in features, while also revealing how it has a very different internal
structure and design.

The primary goal of this paper is to present some of these issues,
with a limited consideration of some other tools that exist for 
nonlinear least squares calculations. A hope is that we will encourage
others to engage, possibly with us, to find ways to bring improvements.
We note that the referees' comments have been most helpful to focus
the discussion here in ways that illustrate the best aspects of open
source software development. We believe it is desirable to consolidate
tools where reasonable so that users have an easier task in choosing
which to use, but such consolidation demands careful and wide-ranging
discussion of options. It is much simpler to create new packages than
to bring together and simplify existing ones.

### Tools considered

Besides the base-R `nls()` function, we consider some packages that 
are available on the CRAN repository.
Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
**minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
general nonlinear least-squares solvers. We will 
not pursue tools in the Bioconductor (@Bioconductor) 
collection, nor those on repositories such as 
GitHub (https://github.com) and Gitlab (https://about.gitlab.com).

@@

?? check wording

@@

`optimx` (@p-optimx) -- This provides 
optimizers that can be applied to minimize a nonlinear function which could be
a nonlinear sum of 
squares. not generally recommended if nonlinear 
least squares programs can be easily used, but provides a check and alternative
solvers.

@@

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by not forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods.

@@

Let us specify in R the three model formulas and set some starting
values for parameters. These starting points are not equivalent and
are deliberately crude choices. Workers performing many calculations 
of a similar nature should try to provide good starting points to reduce
computation time and avoid finding a false solution.

@@

However, `nls()` also offers the option of using the "port" library algorithm
choice (@FoxHallSchryer1978). There is a warning in the `nls()` documentation that
this "appears unfinished", but it is successful in solving all three cases.

```{r ex02nlsport}
unls1p<-try(nls(formula=frmu, start=stu1, data=weeddf, algorithm="port"))
pnls(unls1p)
snls1p<-try(nls(formula=frms, start=sts1, data=weeddf, algorithm="port"))
pnls(snls1p)
tnls1p<-try(nls(formula=frmt, start=stt1, data=weeddf, algorithm="port"))
pnls(tnls1p)
```

@@

Though we have found solutions, the Jacobian is essentially singular as 
shown by its singular
values. Note that these are **displayed** by package `nlsr` in a single column in
the output to provide a compact layout, but the values do **not** correspond to 
the individual parameters in 
whose row they appear; they are a property of the whole problem.

@@

```{r ex02gslnls}
library(gslnls)
ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
pnls(ugslnls1) # to get sum of squares
sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
pnls(sgslnls1) # Use summary() to get display
tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
pnls(tgslnls1) 
## Other algorithms
## tgslnls1a<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))
## pnls(tgslnls1a) 
## tgslnls1d<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))
## pnls(tgslnls1d) 
## tgslnls1dd<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))
## pnls(tgslnls1dd) 
## tgslnls1s2<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))
## pnls(tgslnls1s2) 
```

@@

`nlsr::nlxb()` uses `print()` to output standard errors and 
singular values of the Jacobian (for diagnostic purposes). By contrast, 
`minpack.lm::nlsLM()` and `nls()`
use `summary()`, which does not display the sum of squares, while `print()` 
gives the sum of squares, but not the standard error of the residuals.
As pointed out by one of the referees, such behaviour is consistent
with other modeling functions such as `lm()`.

The importance of the  singular values is to allow us to gauge how "nearly singular" the 
Jacobian is at the solution, and the ratio of the smallest to largest of the 
singular values is a simple but effective measure. The ratios are
4.6641e-05 for Logistic3U, 0.021022 for Logistic3S, and 0.001055 for
Logistic3T, so Logistic3S is the "least singular".

@@

We see that there are differences in detail, but the more important result is that
two out of three singular values are essentially 0. Our Jacobian is singular, and no 
method of the Gauss-Newton type should be able to continue. Indeed, from the parameters
reported at this saddle point, `nlsr::nlxb()` cannot proceed. We also note that
changing the start to
```
stt2<-c(Asym=1, xmid=1, scal=2)
```
allows both `nlsLM` and `gsl_nls` to find the minimum.

@@

We consider the code inside package `nlsr` to be competently coded, with
the advantage over `nls()` of the Levenberg-Marquardt
stabilization in common with `minpack.lm` and `glsnls`. However, our principal 
message is that we obtain diagnostic information in the form of the singular values
of the Jacobian that allow us to try alternative starts, algorithms or programs in 
cases where there are indications of early termination.

@@

Let us write a slightly modified form of the nonlinear least squares problem:

$$ S(p) = \sum_{i=1}^m { r_i(p)^2 }$$

R users are likely to think almost always that the functions $r_i$ are all of
the same form, differing in the details of the data they use. However, the
famous Rosenbrock banana-shaped valley (@Rosenbrock1960, 
also https://en.wikipedia.org/wiki/Rosenbrock_function) has 
$$ r_1(p) = (1 - p_1)$$
$$ r_2(p) = 10 (p_2 - {p_1}^2)$$

Thus there can be problems where the idea of a "model" or a "prediction"
makes no sense, though the nonlinear least squares problem is still
well-posed. In our code, we still refer to the functions as "residuals",
though this is purely a convenience.

@@

We note that the analytic expressions for the Jacobian ("gradient") in the `SSLogis` 
function and others save quantities by default in "hidden" variables, i.e., with 
names preceded by ".".  A referee has indicated this behaviour is to avoid name 
conflicts with user quantities. Hidden variables are not displayed by the `ls()` 
command, so users need to remember to use `ls(all.names=TRUE)` to see which quantities
are defined, for example, in developing their own selfStart model via copy and edit.

Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
worth considering as another way to find starting parameters, but we know of no
R capability of this type.

@@

One of the referees pointed out that nlsr::nlfb() is quite slow on a
"large" case of the Penalty 1 test problem of @More1981TUO. We agree. 
However, we would suggest that problems in large numbers of parameters
will be problematic for Gauss-Newton methods because they create large
computational matrices for the Jacobian or its inner product. The method
"Rcgmin" from package `optimx` (@p-optimx) can be orders of magnitude
faster, but is entirely written in R. Note that "Rcgmin" is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter.

@@

`nls()` can be called without specifying the `data` argument. In this case, it will
search in the available environments (i.e., workspaces) for suitable data objects. 
We do not like this approach, but it is "the R way". R allows users to leave many 
objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
suppress saving this workspace (`.RData`) on exit; otherwise, any such file in 
the path will be loaded on startup. R users in our
acquaintance avoid saving the workspace because of lurking data and
functions that may cause unwanted results.
